---
title: "Identifying Major Players in Endangered Wildlife Trade"
author: '<span class="meta">Lemuel Kumarga</span>'
date: '<span class="meta">May 2018</span>'
always_allow_html: yes
knit: (function(inputFile, encoding) { source("shared/knit.R"); knitter(inputFile, encoding) })
---

```{r echo=FALSE, warning=FALSE, results='hide'}
packages <- c("knitr")
tmp <- lapply(packages, library, character.only = TRUE)

# Set images globally
opts_chunk$set(fig.align='center', fig.height = 4, fig.width = 7)
read_chunk("main.R")
  
```

<style>
/* Add custom css for this particular portfolio */
.players-container {
  display: flex; 
  flex-wrap: wrap; 
  justify-content: space-around
}

a.players {
    margin: 1em;
}

a.players img {
    width: 100px;
    min-width: 100px;
    height: 100px;
    min-height: 100px;
    opacity: 1.0;
    filter: saturate(0%);
    -webkit-filter: saturate(0%);
}

a.players:hover img {
    opacity: 1.0;
    filter: saturate(80%);
    -webkit-filter: saturate(80%);
}

.players-container .popover {
    max-width: 250px;
}
</style>

## Problem Description

Since the Stone Age, mankind has turned to nature for food, commerce and companionship. Our impact back then was minimal as resources were consumed at sustainable rates. However, over the past few decades, nature is increasingly under threat from overconsumption and excess capitalism. Our planet is facing multiple challenges on the environmental front, ranging from global warming to forest degradation. 

One corrosive impact of our greed is the shrinking wildlife diversity and population. In 2000, IUCN counted a total of around <a href="https://portals.iucn.org/library/sites/library/files/documents/RL-2000-001.pdf" target="_blank">10,000 endangered species</a>. By 2017, this number has doubled to <a href="https://www.statista.com/statistics/269910/red-list-endangered-animals-2010-and-2000/" target="_blank">24,431 species</a>. In the face of such overwhelming statistics, it is imperative for us to take action and protect these species from extinction. In this project, we will aim to contribute by using data from both <a href="https://www.cites.org/eng/disc/what.php" target="_blank">CITES (Convention on Internation Trade in Endangered Species of Wild Fauna and Flora)</a> and <a href="http://www.iucnredlist.org/about/introduction" target="_blank">IUCN</a> <span class="hl">to identify the major players behind endangered wildlife trade</span>.

To skip the methodology and proceed straight to the results, please click <a href="#summary-of-results">here</a>.

## Preliminaries

First load the necessary packages for this exercise.
```{r}
<<init>>

si <- sessionInfo()
base_pkg_str <- paste0("Base Packages: ",paste(si[["basePkgs"]], collapse=", "))
attached_pkg_str <- paste0("Attached Packages: ",paste(names(si[["otherPkgs"]]), collapse=", "))
cat(paste0(base_pkg_str,"\n",attached_pkg_str))
```

## About the Data

We will be using wildlife trade data from CITES for the period of 2001 to 2015, with the following caveats:

* 2016 and 2017 were excluded due to data lag in the year of analysis (2018). (See <a href="https://trade.cites.org/cites_trade_guidelines/en-CITES_Trade_Database_Guide.pdf" target="_blank">Section 1.2.2 of the guide</a> for more details.)
* Analysis will be restricted to trades whose sources originated from the wild.

Listed below is an overview of the CITES data:
```{r}
<<data-about>>
  
pander(cols_summary, caption='Wildlife Trade Data - For more info, please visit <a href="https://trade.cites.org/" target="_blank">CITES Trade Database</a>')
```

From the <a href="https://trade.cites.org/cites_trade_guidelines/en-CITES_Trade_Database_Guide.pdf" target="_blank">guide</a> and the above summary, we know that:

1. Each row corresponds to the <span class="hl">total trade</span> between two countries for a particular species at a particular term. This is contrary to popular belief that each row corresponds to one shipment. (See Section 3.1 for more details.)
2. Terms are <span class="hl">heterogenous</span>. For example, some quantities correspond to bodies, while others correspond to feathers.
3. Units are also <span class="hl">varied</span>. Quantities can be quoted as distinct counts (i.e. blank unit), or in terms of weight/volume/qualitative units.
4. Not all the taxonomies are complete. Some rows have missing <span class="hl">Class, Order, Family and/or Genus</span>. It is important for us to fill in these taxonomies to determine each species' trades.
5. Not all animals in the data are endangered. For example, the <a href="https://en.wikipedia.org/wiki/White-tailed_eagle" target="_blank">white-tailed eagle (Haliaeetus albicilla)</a> is specified as <span class="hl">Least Concern</span> on the IUCN Red List.

As can be seen, some pre-processing would be required before our analysis can proceed. In particular, (2) and (3) need to be standardized to allow comparison across species.

## Pre-Processing

To skip the pre-processing, please click <a href="#exploration">here</a>.

Firstly, let us exclude plants from the scope of this analysis:

```{r}
animal_remove <- tictoc(function() { nrow(dataset) },
                        function(old, new) {
                          cat(sprintf("%s rows removed (%i%% of total)",comma(old - new),floor((old - new)/old * 100)))  
                        })
animal_remove$tic()

<<pp-animals>>

animal_remove$toc()
```

### Standardizing the Terms

Next, we need to standardize all the terms below into universal <span class="hl">animal units</span>:

```{r}
<<pp-terms-overview>>

output_str <- "List of Terms:\n" %>%
               paste0(paste0(unique(dataset$Term), collapse=", "),"\n\n") %>%
               paste0("List of Units:\n") %>%
               paste0(paste0(unique((dataset %>% 
                       mutate(Unit = ifelse(Unit == "","count",Unit)))$Unit), 
                       collapse=", "))

cat(output_str)
```

#### Scientific Units
The first thing to note is that not all <span class="hl">units</span> are SI (e.g. <span class="hl">cm, g and litres</span>). This is relatively straightforward to fix:

```{r}
terms_remove <- tictoc(function() { nrow(dataset %>% select(Term, Unit) %>% unique()) },
                        function(old, new) {
                           cat(sprintf("%s term-unit pair%s remaining (%i%% of total removed)",
                              comma(new),
                              ifelse(new == 1,"","s"),
                              floor((old - new)/old * 100))) 
                        },
                       tic_on_toc = TRUE)
terms_remove$tic()

<<pp-to-si>>

terms_remove$toc()
```

#### One Term, One Unit
Another issue is that one term can be recorded in different units. Consider <span class="hl">bodies</span>, which were recorded in counts, kilograms and even metres:

```{r}
<<pp-term-unit-summary>>
  
output_tbl <- term_unit_counts %>%
              filter(Term == "bodies")

pander(output_tbl)
```

Ideally, we would convert kilograms of bodies into actual count by identifying the species' weights. However, such methods are time-intensive. Therefore, we need to propose a conversion rule that is more manageable and yet still relatively accurate.

To standardize units for each term, we will first choose the target unit to convert to. This can be done by identifying the unit with the highest number of records.

```{r}
<<pp-term-unit-target>>
```

The tricky part comes in when we convert a quantity from a non-target unit to the target unit. During this step, we will assume that, <span class="hl">for each species, the median quantity traded in animal units is constant regardless of the terms/units</span>. In other words, if the median quantity of elephant bodies is 2, and the median quantity of elephant bodies in kgs is 14,000, then the 2 bodies are equivalent to 14,000 kgs.

This assumption allows us to convert quantities to the target unit using the following equation:

\[ q_{t} = \frac{q_{nt}}{m_{nt}} \times m_{t} \]

where <br>
\(q_{t}\) corresponds to the quantity traded (in target unit), <br>
\(q_{nt}\) corresponds to the quantity traded (in non-target unit), <br>
\(m_{nt}\) corresponds to the median quantity traded (in non-target unit), and <br>
\(m_{t}\) corresponds to the  median quantity traded (in target unit).

##### The Roll-Up Median Approach

One potential challenge with the above approach is the lack of data points for each species, term and unit triplet. To illustrate, consider the African Elephant (Loxodonta africana) trades below:

```{r}
output_tbl <- dataset %>% 
              filter(Taxon == "Loxodonta africana") %>%
              left_join(target_unit, by=c("Term","Unit")) %>%
              filter(is.na(NumberOfUnits)) %>%
              group_by(Taxon, Term, Unit) %>%
              summarise(Records = n())

pander(head(output_tbl,5))
```

There is only a <span class="hl">single</span> data point whose term and unit is "bodies" and "kg"! Such minute sample size is not sufficient to calculate the median. To obtain larger sample sizes, a roll-up approach is required:

```{r}
# For documentation, please visit http://rich-iannone.github.io/DiagrammeR/graphviz_and_mermaid.html
grViz(paste0("
  digraph RollUp {

    # Default Specs
    graph [compound = true, nodesep = .5, ranksep = .25]
    node [fontname = '",`@f`,"', fontsize = 14, fontcolor = '",`@c`(bg),"', penwidth=0, color='",`@c`(ltxt),"', style=filled]
    edge [fontname = '",`@f`,"', fontcolor = '",`@c`(ltxt),"', color='",`@c`(ltxt),"']
    
    # Input Specs
    Inp [fillcolor = '",`@c`(ltxt),"', label = '(Species s, Term t, Unit u)', shape = rectangle]

    # Conclude Specs
    node [shape = oval]
    Species_Y [label = 'Use the median quantity\nof Species s, Term t and Unit u.', fillcolor = '",`@c`(1),"']
    Genus_Y [label = 'Use the median quantity\nof Genus g, Term t and Unit u.', fillcolor = '",`@c`(2),"']
    Family_Y [label = 'Use the median quantity\nof Family f, Term t and Unit u.', fillcolor = '",`@c`(3),"']
    Order_Y [label = 'Use the median quantity\nof Order o, Term t and Unit u.', fillcolor = '",`@c`(4),"']
    Class_Y [label = 'Use the median quantity\nof Class c, Term t and Unit u.', fillcolor = '",`@c`(5),"']
    Kingdom_Y [label = 'Use the median quantity\nof Term t and Unit u (across all).', fillcolor = '",`@c`(txt),"']

    # Trigger Specs
    node [shape = diamond, fillcolor = '",`@c`(bg),"', fontcolor = '",`@c`(txt),"', penwidth = 1]
    Species_T [label = 'Does the Species s\nhave >=10 records\nfor the term t and unit u?']
    { rank = same; Species_T, Species_Y }
    Genus_T [label = 'Does g (the Genus of s)\nhave >=10 records\nfor the term t and unit u?']
    { rank = same; Genus_T, Genus_Y }
    Family_T [label = 'Does f (the Family of g)\nhave >=10 records\nfor the term t and unit u?']
    { rank = same; Family_T, Family_Y }
    Order_T [label = 'Does o (the Order of f)\nhave >=10 records\nfor the term t and unit u?']
    { rank = same; Order_T, Order_Y }
    Class_T [label = 'Does c (the Class of o)\nhave any record\nfor the term t and unit u?']
    { rank = same; Class_T, Class_Y }

    # Yes edges
    Inp -> Species_T
    edge [arrowhead = 'box', label = 'Yes']
    Species_T -> Species_Y
    Genus_T -> Genus_Y
    Family_T -> Family_Y
    Order_T -> Order_Y
    Class_T -> Class_Y
    
    # No edges
    edge [label = '     No']
    Species_T -> Genus_T
    Genus_T -> Family_T
    Family_T -> Order_T
    Order_T -> Class_T
    Class_T -> Kingdom_Y

  }
"), height=700)
```

To sum up the flowchart above, we first count the number of records under the given species, term and unit. If the records are insufficient, we then "roll-up" and find the number of records for the genus associated with the species. This process continues until we have sufficient data points to calculate the median.

##### Managing Outliers

Another potential challenge of converting around the median is that outliers may be extremely high, skewing the aggregate results towards certain records. To remedy this, we will floor the scaled quantity (\( q_{nt} / m_{nt} \)) of each record to the 5th percentile and cap it to the 95th percentile.

Using the following process, we can now convert all the terms to their standardized units.
```{r}
<<pp-term-unit-final>>

terms_remove$toc()
```

#### Well-Defined Terms

Let us now <span class="hl">convert each term to the universal animal unit</span>. Through observation, the terms can be differentiated into well-defined or ambiguous. A term is well-defined if each animal has the same number of them. One good example would be tusk, since 2 of them are always the equivalent of 1 animal. 

Listed below are the terms we describe as well-defined:

```{r}
<<pp-term-well-defined-overview>>

output_str <- "Well-Defined Terms [with Number Per Live Animal]:\n"
wdt <- well_defined_terms %>%
       mutate(outputString = paste0(Term," [", perAnimal, "]"))
output_str <- output_str %>%
              paste0(paste0(wdt$outputString,collapse=", "))
  
cat(output_str)
```

To convert a well-defined term, we simply need to divide the quantity by the number of terms per animal:

```{r}
<<pp-term-well-defined-convert>>

terms_remove$toc()
```

#### Ambiguous Terms

The final piece of the jigsaw is to convert ambiguous terms. These terms, by nature of their names, are somewhat vague, and there exists no clear association between them and the number of animals. For example, how many ivory pieces make up an elephant? Depending on the size of the pieces, the numbers would differ record by record.

To simplify the conversion process, we will revisit our previous assumption when standardizing units of a term. Assuming that <span class="hl">across each species, the median quantity traded in live animal units is the same</span>, ambiguous terms can be converted using the following equation:

\[ q_{animal} = \frac{q_{at}}{m_{at}} \times m_{animal} \]

where <br>
\(q_{animal}\) corresponds to the quantity of animals traded, <br>
\(q_{at}\) corresponds to the quantity traded (in ambiguous terms), <br>
\(m_{at}\) corresponds to the median quantity traded (in ambiguous terms), and <br>
\(m_{animal}\) corresponds to the median quantity of animals traded.

As before, medians are obtained through the <a href="#the-roll-up-median-approach">roll-up approach</a>, and outliers are managed by <a href="#managing-outliers">flooring and capping at the 5th and 95th percentile</a> respectively.

With this final step, we have reduced over 200 distinct term-unit pairs into a single animal unit!

```{r}
<<pp-term-ambiguous-convert>>

terms_remove$toc()
```

### Completing the Taxonomy

One of the problems with incomplete taxonomies is that identical species have different identities. For example, it is possible that an animal was recorded with its species taxonomy during importing, but only with an order during exporting. In such circumstances, we might misinterpret the former as being consumed, while the latter as being captured locally. 

To correct for the incomplete taxonomies, we will use the method below:

```{r}
grViz(paste0("
  digraph CompleteTax {

    # Default Specs
    graph [compound = true, nodesep = .5, ranksep = .25]
    node [fontname = '",`@f`,"', fontsize = 14, fontcolor = '",`@c`(bg),"', penwidth=0, color='",`@c`(ltxt),"', style=filled]
    edge [fontname = '",`@f`,"', fontcolor = '",`@c`(ltxt),"', color='",`@c`(ltxt),"']
    
    # Input Specs
    Inp [fillcolor = '",`@c`(ltxt),"', label = 'Incomplete Taxonomy Trades from Country A to B\nfor Class c and Order o', shape = rectangle]

    # Conclude Specs
    node [shape = oval]
    Route_Y [label = 'Divide the trades proportionally\namong Class c, Order o\n trades from A to B.', fillcolor = '",`@c`(1),"']
    Export_Y [label = 'Divide the trades proportionally\namong Class c, Order o\n trades from A.', fillcolor = '",`@c`(2),"']
    Import_Y [label = 'Divide the trades proportionally\namong Class c, Order o\n trades to B.', fillcolor = '",`@c`(3),"']
    All_Y [label = 'Divide the trades proportionally\namong all Class c, Order o\n trades.', fillcolor = '",`@c`(4),"']

    # Trigger Specs
    node [shape = diamond, fillcolor = '",`@c`(bg),"', fontcolor = '",`@c`(txt),"', penwidth = 1]
    Route_T [label = 'Does A to B have\nany complete taxonomy trades\ncorresponding to Class c and Order o?']
    { rank = same; Route_T, Route_Y }
    Export_T [label = 'Does A have\nany complete taxonomy exports\ncorresponding to Class c and Order o?']
    { rank = same; Export_T, Export_Y }
    Import_T [label = 'Does B have\nany complete taxonomy imports\ncorresponding to Class c and Order o?']
    { rank = same; Import_T, Import_Y }

    # Yes edges
    Inp -> Route_T
    edge [arrowhead = 'box', label = 'Yes']
    Route_T -> Route_Y
    Export_T -> Export_Y
    Import_T -> Import_Y
    
    # No edges
    edge [label = '     No']
    Route_T -> Export_T
    Export_T -> Import_T
    Import_T -> All_Y

  }
"), height=500)
```

In summary, the methodology above distributes incomplete taxonomy trades to their related species. The approach first analyzes whether there are species of similar natures undergoing the same trade route. If there are, the incomplete taxonomy trades will be distributed amongst the species at existing ratios. If there are none, we subsequently "roll-up" and consider the species exported by the Exporter. This process continues until all the incomplete taxonomy trades are attributed to their respective species.

```{r}
taxon_remove <- tictoc(function() { nrow(dataset %>% filter((Taxon == Class | grepl("spp.",Taxon) | Genus == ""))) },
                       function(old,new) {
                         cat(sprintf("%s incomplete taxonomies converted",comma(old-new)))
                       })
taxon_remove$tic()

<<pp-taxon>>
  
taxon_remove$toc()
```

### Restricting to Endangered Species

A species's endangered status is located in the IUCN Red List database. Due to licensing restrictions, the data could not be uploaded for public view. However, one is free to create an account and download the csv from <a href="http://www.iucnredlist.org/search/saved?id=90695" target="_blank">here</a>.

By integrating the IUCN data, we can filter non-endangered species out of the dataset.
```{r}
animal_remove$tic()

<<pp-endanger>>

animal_remove$toc()
```

## Exploration

#### Evolution Across Time

The interactive chart below shows how wildlife trade evolves across time:

```{r}
<<exp-time>>

streamgraph_plot
```

Based on the above, we know that as of 2011, species whose statuses are <span class="hl red-text">Critical</span> account for <span class="hl red-text">`r sprintf("%.0f%%",(trades_by_time %>% filter(Year == 2011) %>% mutate(pct = total_trades/sum(total_trades)) %>% filter(IUCNLabel == "Critical"))$pct * 100)`</span> of the total trades. By 2015, however, these critical species account for <span class="hl red-text">`r sprintf("%.0f%%",(trades_by_time %>% filter(Year == 2015) %>% mutate(pct = total_trades/sum(total_trades)) %>% filter(IUCNLabel == "Critical"))$pct * 100)`</span> of the total trades. This is disconcerting as the species closest to extinction are the ones being captured the most.

#### Reasons for Trading

The waffle chart below highlights the purposes behind endangered wildlife trade:

```{r, fig.asp=0.35}
<<exp-purpose>>

waffle_plot
```

Shockingly, trading due to <span class="hl red-text">commercial</span> reasons account for over <span class="hl red-text">94%</span> of all trades! In other words, for every 1 animal traded for <span class="hl green-text">conservation</span> purposes, 188 trades are being conducted for profit-making. It seems that conservationists are having difficulties catching up with excess capitalism.

#### Most Traded Species

The sunburst plot below allows us to find out which animal class, and which species within each class are traded the most:

```{r}
<<exp-species>>

# Centerize the sunburst plot
htmlwidgets::onRender(
    sunburst_plot,
    "
    function(el, x) {
    $('.sunburst-chart').css('width', 'initial');
    $('.sunburst-chart').css('height', 'initial');
    $('.sunburst-chart').css('left', '50%');
    $('.sunburst-chart').css('transform', 'translate(-50%, 0)');
    }
    "
)

```

By scrolling through the chart above, we know that <span class="hl yellow-text">Anthozoa</span> accounts for close to <span class="hl yellow-text">33%</span> of the total trades. Some examples of species belonging to this class are soft and hard corals. The well-publicized <span class="hl blue-text">African Elephant</span>, belonging to the <span class="hl blue-text">Mammalia</span> class, interestingly accounts for less than 0.4% of the total trades.

#### Demand Hotspots

Using export and import data, we can estimate the consumption of wildlife in each country. If a country imports more than it exports, then it is reasonable to assume that the difference is consumed locally.

```{r}
<<exp-imports>>

leaflet_import_plot
```

By utilizing net imports as a proxy, we determined countries with the highest wildlife demand. Surprisingly, <span class="hl red-text">China</span> is ranked third, while the <span class="hl red-text">United States</span> has around 3 times more demand than the second-ranked country, <span class="hl red-text">Japan</span>! <span class="hl red-text">Korea</span> and <span class="hl red-text">Hong Kong</span> make up the top 5.

It remains to be seen whether the huge gap between United States and the other top 4 countries is due to actual trading or higher diligence in reporting.

#### Wildlife Supply

Using the opposite logic as demand, if a country exports more than it imports, then we assume that the difference is captured locally.

```{r}
<<exp-exports>>

leaflet_export_plot
```

Using net exports as a proxy, we can determine the distribution of supply across the globe. As expected, Southeast Asia, Africa and South America are areas where the majority of wildlife is coming from. In particular, <span class="hl purple-text">Indonesia</span> exports the most wildlife, close to 2 times than the second-ranked <span class="hl purple-text">Ecuador</span>. <span class="hl purple-text">Morocco</span>, the highest-ranked nation in Africa, is 5th.

## Identifying the Major Players

For the purposes of this analysis, the major players in the endangered wildlife trade are countries which take part extensively as either suppliers, consumers or intermediaries. In this section, we will aim to:

1. surface players that were otherwise overlooked, and
2. rank each player according to its wildlife trading activities.

[1] will allow us to have a more macro perspective on each country's role towards species degradation, while [2] will help us prioritize which countries to focus on given limited conservation resources.

### Choosing the Algorithm

The most straightforward way to identify major players is by <span class="hl">their net imports/exports</span>. By ordering these metrics, we can potentially find countries supplying and consuming endangered wildlife goods. However, one potential pitfall of such an approach is that the intermediary players get overlooked. Consider the following scenario:

```{r}
grViz(paste0("
  digraph expInt {

    # Default Specs
    graph [compound = true, nodesep = .5, ranksep = .25]
    node [fontname = '",`@f`,"', fontsize = 14, fontcolor = '",`@c`(bg),"', penwidth=0, color='",`@c`(ltxt),"', style=filled, shape = oval]
    edge [fontname = '",`@f`,"', fontcolor = '",`@c`(ltxt),"', color='",`@c`(ltxt),"']
    
    Country_A [label = 'Country\nA', fillcolor = '",`@c`(1),"']
    Country_B [label = 'Country\nB', fillcolor = '",`@c`(2),"']
    Country_C [label = 'Country\nC', fillcolor = '",`@c`(3),"']
    { rank = same; Country_A, Country_B, Country_C }

    Country_A -> Country_B [ label='100' ]
    Country_B -> Country_C [ label='99' ]
    
  }
"), width=309, height=50)
```

In such a scenario, <span class="color-1-text">A</span> will have a net export of 100, B a net import of 1 and <span class="color-3-text">C</span> a net import of 99. Using the net import/export approach, <span class="color-2-text">B</span> is the least important player. However, in actuality, <span class="color-2-text">B</span> should be the most pivotal country as it is involved in all 199 trades. Hence, a better approach is required.

Another way to tackle the issue above is to instead use each country's gross <span class="hl">imports + exports</span>. This approach would work well in the above case, as <span class="color-2-text">B</span> would rightly be identified as the most important player. However, one weakness of such a method would be the assumption that absolute numbers are extremely accurate. In truth, this may not be the case, due to the following reasons:

1. There may be reporting errors in the CITES database, and
2. There could be conversion errors incurred in the <a href="#pre-processing">Pre-Processing</a> stage.

Moreover, consider the following scenario: 

```{r}
grViz(paste0("
  digraph expInt {

    # Default Specs
    graph [compound = true, nodesep = .5, ranksep = .25]
    node [fontname = '",`@f`,"', fontsize = 14, fontcolor = '",`@c`(bg),"', penwidth=0, color='",`@c`(ltxt),"', style=filled, shape = oval]
    edge [fontname = '",`@f`,"', fontcolor = '",`@c`(ltxt),"', color='",`@c`(ltxt),"']
    
    Country_A [label = 'Country\nA', fillcolor = '",`@c`(1),"']
    Country_B [label = 'Country\nB', fillcolor = '",`@c`(2),"']
    Country_C [label = 'Country\nC', fillcolor = '",`@c`(3),"']
    Country_D [label = 'Country\nD', fillcolor = '",`@c`(4),"']
    Country_E [label = 'Country\nE', fillcolor = '",`@c`(5),"']

    Country_A -> Country_B [ label='  100' ]
    Country_B -> Country_C [ label='  99' ]
    Country_A -> Country_D [ label='  200' ]
    Country_D -> Country_E [ label='  99' ]
  }
"), width=173, height=200)
```

Using the imports + exports approach, we conclude that both <span class="color-3-text">C</span> and <span class="color-5-text">E</span> have the same degree of involvement. However, if the numbers presented are only <span class="hl">roughly</span> accurate, it would be reasonable to assume that <span class="color-5-text">E</span> is more involved as it is linked to a bigger intermediary in <span class="color-4-text">D</span>. (This same argument could be made for drug trafficking, where the closer you are to a drug cartel, the more likely you are to be a major player.)

Fortunately, we can resolve these two issues using <a class="hl" href="https://en.wikipedia.org/wiki/PageRank">Google's PageRank</a> algorithm. The algorithm is famous for ranking a website's quality based on:

i. the number of links pointing to the website, and 
ii. who is pointing to the website.

This algorithm can be leveraged for our analysis to assess the involvement of a country by:

i. the number of trades the country is involved in, and
ii. which other nations the country is involved with.

### Modeling Wildlife Trade

To run the PageRank algorithm, we need to model the wildlife trade as a network, with the countries represented as <span class="hl">vertices</span>. Two countries are connected by an <span class="hl">edge</span> if the two trade with one another. The degree of wildlife trading between two countries defines the <span>weight</span> of the edge. For the purposes of this study, we will use the sum of imports and exports as the edges' weights.

Using the above definitions, we are now equipped to model the wildlife trade for the period of 2001 to 2015:

```{r,fig.asp=1.05}
<<model-graphs>>
  
edge_bundle_plot
```

The <a href="https://www.r-graph-gallery.com/hierarchical-edge-bundling/" target="_blank">hierarchical edge bundle</a> plot shows the degree of involvement of each country. The larger the circle, the greater the total imports and exports of the country. Unsurprisingly, <span class="hl blue-text">United States</span>,  <span class="hl orange-text">China</span> and <span class="hl orange-text">Indonesia</span> have the highest total imports and exports.

Other than that, we can also identify the most popular trade routes. The larger the wildlife trading, the higher the opacity of that edge. The chart above surfaces some overlooked routes, such as <span class="hl blue-text">Ecuador</span> to <span class="hl orange-text">China</span>. 

### The PageRank Algorithm

With the network constructed, we can proceed to run the PageRank algorithm:

```{r}
<<model-pagerank>>

pander(vertices %>% arrange(desc(PRANK)) %>% head(5) %>%
       mutate(Country=NAME, Imports_And_Exports = comma(VALUE), PageRank = percent(PRANK)) %>% 
         select(Country, Imports_And_Exports, PageRank),
       caption="The Top 5 Major Players According to PageRank")
```

Before continuing, it is important for us to understand what the PageRank number represents. When used for assessing the quality of websites, this number refers to a person's <span class="hl">probability of viewing the website</span>. The higher the PageRank probability, the more likely the person will view the website, and hence the higher the quality of the site.

Similarly, in the widlife trading context, the PageRank of a country A is the <span class="hl">probability that A is responsible for a wildlife trade</span>. The higher the PageRank of a country, the more wildlife trades are attributed to it. In other words, if there are 1 million trades worldwide, <span class="hl">United States</span> would account for approximately <span class="hl">`r round((vertices %>% filter(index == "US"))$PRANK,3)` million</span> trades. By identifying countries with the highest PageRank, we can determine who the major players are.

#### Comparing With Imports + Exports Approach

To comprehend further, let us first compare PageRank with Imports + Exports:

```{r, fig.asp=0.5}
<<model-pagerank-hist>>

hist_plot
```

The plot above shows the distribution of <span class="hl blue-text">PageRank</span> and <span class="hl green-text">Imports + Export</span>. The latter distribution has a much gentler slope, implying that wildlife trade responsibilites are more "spread-out" across countries. <span class="hl">The PageRank algorithm, however, extremizes this distribution by compacting "average" countries into a single peak.</span> Only a handful of least and most important nations are left in the tails, allowing us to better distinguish them.

Secondly, let us investigate how the rankings shift as we transition from one approach to another:

```{r, fig.asp=0.7}
<<model-pagerank-rank>>

rank_plot
```

In both rankings, the top 6 countries remain the same. This implies that <span class="hl">imports and exports do play a prominent role in the PageRank algorithm.</span> However, the shifts in rankings also indicate that total trades is not the only factor influencing PageRank. 

To identify other important factors, let us compare two countries that have been promoted and demoted extensively, <span class="hl orange-text">Japan</span> and <span class="hl blue-text">Ecuador</span>:

```{r, fig.asp=1.05}
<<model-pagerank-compare>>

compare_plot
```

The diagram above shows all the countries that either Japan or Ecuador traded with. It is clear that Japan has a greater outreach than Ecuador, even though the latter has a higher total imports + exports. <span class="hl">This degree of outreach caused the algorithm to promote <span class="orange-text">Japan</span> at the expense of <span class="blue-text">Ecuador</span>.</span>

#### Decomposing PageRank

The PageRank algorithm allows us to determine who the major players are. However, each country has different roles to play. Some can be <span class="hl">consumers</span> (countries which fuel the demand of wildlife consumption), <span class="hl">suppliers</span> (countries which capture the wildlife and ship them out) or even <span class="hl">dealers</span> (intermediaries which connect suppliers and consumers). To reduce wildlife trading effectively, tackling different roles will require different strategies. 

<span class="hl">In this section, we will construct a method to identify the top consumers, dealers and suppliers.</span> It is important to note that these roles are not exclusive; a country may well be a supplier of one species and a consumer of another.

To determine a country's role, we will first split the quantity traded of each species into three constituents:

* <span class="hl">Net Imports</span>: The amount of trades being brought into the country. This is the leftover amount when we subtract imports with exports. If exports are more than imports, then there are no net imports.
* <span class="hl">Net Exports</span>: The amount of trades being shipped out of the country. This is equivalent to the amount remaining when we subtract exports with imports. If imports are more than exports, there are no net exports.
* <span class="hl">Transits</span>: These are the trades being brought in and then shipped out of the country. This is equivalent to the minimum of Exports and Imports.

These constitutients are then summed up across species to determine a country's net imports, net exports and transits. Thee 3 values are subsequently used to decompose PageRank via the following approach:

```{r}
grViz(paste0("
  digraph decomPR {

    # Default Specs
    graph [compound = true, nodesep = .5, ranksep = .25]
    node [fontname = '",`@f`,"', fontsize = 14, fontcolor = '",`@c`(bg),"', penwidth=0, color='",`@c`(ltxt),"', style=filled, shape = oval]
    edge [fontname = '",`@f`,"', fontcolor = '",`@c`(ltxt),"', color='",`@c`(ltxt),"']
    
    PageRank [fillcolor = '",`@c`(ltxt),"', shape=diamond]
    ConsumerRank [fillcolor = '",`@c`(red),"']
    DealerRank [fillcolor = '",`@c`(sec),"']
    SupplierRank [fillcolor = '",`@c`(purple),"']

    PageRank -> ConsumerRank [ label=' x Net Imports %' ]
    PageRank-> DealerRank [ label='  x Transit %' ]
    PageRank -> SupplierRank [ label='  x Net Exports %' ]
  }
"), width=405, height=100)
```

where

* <span class="hl red-text">ConsumerRank</span> is the % of trades due to the country's <span class="hl red-text">consumption</span>,
* <span class="hl">DealerRank</span> is the % of trades due to the country <span class="hl">being an intermediary</span>, and
* <span class="hl purple-text">SupplierRank</span> is the % of trades due to the country's <span class="hl purple-text">provision<span>.

By ordering each of these three ranks, we can determine which countries are the top consumers, dealers and suppliers. Summing up each of these metrics would also allow us to determine the major drivers behind wildlife trading:

```{r}
<<model-pagerank-decompose-overview>>

"% Trade Contribution by Roles:\n" %>%
paste0(paste0(sapply(c("CRANK","DRANK","SRANK"), function(r) {  
    sprintf("%ss: %.2f%%", ifelse(r == "CRANK","Consumer", ifelse(r == "DRANK","Dealer","Supplier")),sum(vertices[r]) * 100)
  }), collapse="\n")) %>%
cat()
```

From the above summary, consumption and supply of wildlife goods contribute the most to wildlife exchanges. Intermediaries surprisingly do not play major roles. This suggests that to reduce trading, we are better off focusing on reducing supply and demand (instead of breaking down the channels connecting them).

```{r, fig.asp=0.8, width=8}
<<model-pagerank-decompose-leaf>>

suppressWarnings(print(leaf_plot))
```

The leaf plots above show the roles played by the top 12 countries. <span class="hl blue-text">United States</span>, the biggest player, are mostly consumers, while <span class="hl red-text">Germany</span> is a large dealer. <span class="hl orange-text">China</span> unsurprisingly is both consuming and supplying wildlife resources.

### The Major Players

Listed below are the top consumers, dealers and suppliers detected through the PageRank algorithm.

```{r}
<<model-results>>

pander(players, caption="The Top Suppliers, Dealers and Consumers")
```

To find out more about each country's trading activities, please hover or click on the buttons below:

```{r echo=FALSE, warning=FALSE, results='hide'}
<<results>>
```

#### Consumers

`r get_players((consumers %>% head(5))$index,'@c'(red), type="Net_Imports")`

#### Dealers

`r get_players((dealers %>% head(5))$index,'@c'(heading), type="Transits", disclaimers=list("FR"=c("European Eel"="Due to missing black market exports, France is considered an intermediary stop for European Eels, even though they originate from the country. See Limitations.")))`

#### Suppliers

`r get_players((suppliers %>% head(5))$index,'@c'(purple), type="Net_Exports", disclaimers=list("MX"=c("European Eel"="Due to missing black market imports, European Eels are considered as top providers for Mexico. See Limitations."), "CN"=c("European Eel"="Due to missing black market imports, European Eels are considered as top providers for China. See Limitations.")))`

## Summary of Results

By standardizing terms through <a href="#the-roll-up-median-approach">the roll-up median approach</a> and ranking countries via <a href="#the-pagerank-algorithm">the PageRank algorithm</a>, we have identified the major players in endangered wildlife trade:

* <span class="hl">Consumers</span> are countries which generate the demand for wildlife. It is important to educate citizens in these countries to stop using wildlife products. The top 5 consumers are `r paste0(sprintf('<span class="hl">%s</span>',(consumers %>% head(5))$NAME), collapse = ", ")`.
* <span class="hl">Dealers</span> are countries which connect suppliers to consumers. More stringent measures can be set up in the trading ports to reduce endangered wildlife trading. The top 5 dealers are `r paste0(sprintf('<span class="hl">%s</span>',(dealers %>% head(5))$NAME), collapse = ", ")`.
* <span class="hl">Suppliers</span> are countries that provide wildlife goods. Efforts to protect the nature reserves of these countries should be emphasized. The top 5 suppliers are `r paste0(sprintf('<span class="hl">%s</span>',(suppliers %>% head(5))$NAME), collapse = ", ")`.

## Limitations

The above analysis, though simplified, provides us with a better picture of the wildlife situation across the world. Listed below are some suggestions to improve the accuracy of results:

* <span class="hl">One Side of the Story</span>: This study is based on data reported by governments. As such, many black market trades are missing. It would be interesting to identify these "shadow" trades by verifying the consistency of the CITES data, before integrating them back into this analysis.
* <span class="hl">Wild</span>: Due to size constraints, we only considered trades whose sources are tagged as "Wild". By expanding the scope, additional insights could be unearthed.
* <span class="hl">Species-Specific Fixes</span>: The methods in pre-processing are generalized and favor expediency over accuracy. By fixing each trade according to individual species's characteristics (such as weight), we can obtain more accurate transformations of the terms into animal counts.

## References

CITES Database 2017. <https://www.cites.org>. Downloaded on 24 April 2018.

IUCN 2017. The IUCN Red List of Threatened Species. Version 2017-3. <http://www.iucnredlist.org>. Downloaded on 25 April 2018.

