---
title: "Identifying Major Players in Endangered Wildlife Trade"
author: '<span class="meta">Lemuel Kumarga</span>'
date: '<span class="meta">May 2018</span>'
always_allow_html: yes
knit: (function(inputFile, encoding) { source("shared/knit.R"); knitter(inputFile, encoding) })
---

```{r echo=FALSE, warning=FALSE, results='hide'}
packages <- c("knitr")
tmp <- lapply(packages, library, character.only = TRUE)

# Set images globally
opts_chunk$set(fig.align='center', fig.height = 4, fig.width = 7)
read_chunk("main.R")
  
```

<style>
/* Add custom css for this particular portfolio */
a.players {
    margin: 1em;
}

a.players img {
    width: 100px;
    min-width: 100px;
    height: 100px;
    min-height: 100px;
    opacity: 1.0;
    filter: saturate(0%);
    -webkit-filter: saturate(0%);
}

a.players:hover img {
    opacity: 1.0;
    filter: saturate(80%);
    -webkit-filter: saturate(80%);
}
</style>

## Problem Description

Since the Stone Age, mankind has turned to nature for food, commerce and companionship. Our impact back then was minimal as resources were consumed at sustainable rates. However, over the past few decades, nature is increasingly under threat from overconsumption and excess capitalism. Our planet is facing multiple challenges on the environmental front, ranging from global warming to forest degradation. 

One corrosive impact of our greed is the shrinking wildlife diversity and population. In 2000, IUCN counted a total of around <a href="https://portals.iucn.org/library/sites/library/files/documents/RL-2000-001.pdf" target="_blank">10,000 endangered species</a>. By 2017, this number has doubled to <a href="https://www.statista.com/statistics/269910/red-list-endangered-animals-2010-and-2000/" target="_blank">24,431 species</a>. In the face of such overwhelming statistics, it is imperative for us to take action and protect these species from extinction. In this project, we will aim to contribute by using data from both <a href="https://www.cites.org/eng/disc/what.php" target="_blank">CITES (Convention on Internation Trade in Endangered Species of Wild Fauna and Flora)</a> and <a href="http://www.iucnredlist.org/about/introduction" target="_blank">IUCN</a> <span class="hl">to identify the major players behind endangered wildlife trade</span>.

To skip the methodology and proceed straight to the results, please click <a href="#summary-of-results">here</a>.

## Preliminaries

First load the necessary packages for this exercise.
```{r}
<<init>>

si <- sessionInfo()
base_pkg_str <- paste0("Base Packages: ",paste(si[["basePkgs"]], collapse=", "))
attached_pkg_str <- paste0("Attached Packages: ",paste(names(si[["otherPkgs"]]), collapse=", "))
cat(paste0(base_pkg_str,"\n",attached_pkg_str))
```

## About the Data

We will be using wildlife trade data from CITES for the period of 2001 to 2015, with the following caveats:

* 2016 and 2017 were excluded due to data lag in the year of analysis (2018). (See <a href="https://trade.cites.org/cites_trade_guidelines/en-CITES_Trade_Database_Guide.pdf" target="_blank">Section 1.2.2 of the guide</a> for more details.)
* Analysis will be restricted to trades whose sources originated from the wild.

Listed below is an overview of the CITES data:
```{r}
<<data-overview>>
  
pander(cols_summary, caption='Wildlife Trade Data - For more info, please visit <a href="https://trade.cites.org/" target="_blank">CITES Trade Database</a>')
```

From the <a href="https://trade.cites.org/cites_trade_guidelines/en-CITES_Trade_Database_Guide.pdf" target="_blank">guide</a> and the above summary, we know that:

1. Each row corresponds to the <span class="hl">total trade</span> between two countries for a particular species at a particular term. This is contrary to popular belief that each row corresponds to one shipment. (See Section 3.1 for more details.)
2. Terms are <span class="hl">heterogenous</span>. For example, some quantities correspond to bodies, while others correspond to feathers.
3. Units are also <span class="hl">varied</span>. Quantities can be quoted as distinct counts (i.e. blank unit), or in terms of weight/volume/qualitative units.
4. Not all the taxonomies are complete. Some rows have missing <span class="hl">Family and Genus</span>. It is important for us to devise a method to fill in these taxonomies so as to determine the quantity traded of each species.
5. Not all animals in the data are endangered. For example, the <a href="https://en.wikipedia.org/wiki/White-tailed_eagle" target="_blank">white-tailed eagle (Haliaeetus albicilla)</a> is specified as <span class="hl">Least Concern</span> on the IUCN Red List.

As can be seen, some pre-processing would be required before our analysis can proceed. In particular, (2) and (3) need to be standardized to allow comparison across species.

## Pre-Processing

To skip the pre-processing, please click <a href="#exploration">here</a>.

In the data, taxonomies with blank <span class="hl">Classes</span> are plants. For the purposes of this analysis, these records (or rows) will be excluded from the dataset.

```{r}
pre_clean <- nrow(dataset)
post_clean_str <- function () {
  post_clean <- nrow(dataset)
  cat(sprintf("%i rows removed (%i%% of total)",pre_clean - post_clean,floor((pre_clean - post_clean)/pre_clean * 100)))  
}

<<pp-animals>>

post_clean_str()
```

### Standardizing the Terms

After limiting the dataset to only animals, we need to standardize all the terms below into universal <span class="hl">animal units</span>:

```{r}
<<pp-terms-overview>>

output_str <- "List of Terms:\n" %>%
               paste0(paste0(unique(dataset$Term), collapse=", "),"\n\n") %>%
               paste0("List of Units:\n") %>%
               paste0(paste0(unique((dataset %>% 
                       mutate(Unit = ifelse(Unit == "","count",Unit)))$Unit), 
                       collapse=", "))

cat(output_str)
```

#### Scientific Units
The first thing to note is that not all of the <span class="hl">units</span> are SI (e.g. <span class="hl">cm, g and litres</span>). This is relatively straightforward to fix:

```{r}
pre_clean <- nrow(dataset %>% group_by(Term, Unit) %>% summarise())
post_clean_str <- function () {
  post_clean <- nrow(dataset %>% group_by(Term, Unit) %>% summarise())
  cat(sprintf("%i term-unit pair%s remaining (%i%% of total removed)",
              post_clean,
              ifelse(post_clean == 1,"","s"),
              floor((pre_clean - post_clean)/pre_clean * 100)))
}

<<pp-to-si>>

post_clean_str()
```

#### One Term, One Unit
Another issue is that one term can be recorded in different units. Consider the term <span class="hl">bodies</span>, which were recorded in counts, kilograms and even metres:

```{r}
<<pp-term-unit-summary>>
  
output_tbl <- term_unit_counts %>%
              filter(Term == "bodies")

pander(output_tbl)
```

Ideally, we would convert kilograms of bodies into actual count by identifying species' weight. However, such methods are time-intensive, and fall outside the scope of this analysis. Therefore, we need to propose a conversion rule that is more manageable and yet still relatively accurate.

To standardize units for each term, we will first choose the target unit to convert to. This can be done by identifying the unit with the highest number of records.

```{r}
<<pp-term-unit-target>>
```

The tricky part comes in when we attempt to convert a quantity from a non-target unit to the target unit. During this step, we will assume that, <span class="hl">for each species, the median quantity traded in animal units is constant regardless of the terms/units they were specified in</span>. In other words, if the median quantity of elephant bodies traded is 2, and the median quantity of elephant bodies traded in kgs is 14,000, then the 2 elephant bodies are equivalent to 14,000 kgs.

This assumption allows us to convert quantities to the target unit using the following equation:

\[ q_{t} = \frac{q_{nt}}{m_{nt}} \times m_{t} \]

where <br>
\(q_{t}\) corresponds to the quantity traded (in target unit), <br>
\(q_{nt}\) corresponds to the quantity traded (in non-target unit), <br>
\(m_{nt}\) corresponds to the median quantity traded (in non-target unit), and <br>
\(m_{t}\) corresponds to the  median quantity traded (in target unit).

##### The Roll-Up Median Approach

One potential challenge with the above approach is the lack of data points for each species, term and unit triplet. To illustrate, consider the African bush elephant (Loxodonta africana) trades below:

```{r}
output_tbl <- dataset %>% 
              filter(Taxon == "Loxodonta africana") %>%
              left_join(target_unit, by=c("Term","Unit")) %>%
              filter(is.na(NumberOfUnits)) %>%
              group_by(Taxon, Term, Unit) %>%
              summarise(Records = n())

pander(head(output_tbl,5))
```

There is only a <span class="hl">single</span> data point whose term and unit are "bodies" and "kg"! Such minute sample size is not sufficient to calculate the median. To obtain larger sample sizes, a roll-up approach is required:

```{r}
# For documentation, please visit http://rich-iannone.github.io/DiagrammeR/graphviz_and_mermaid.html
grViz(paste0("
  digraph RollUp {

    # Default Specs
    graph [compound = true, nodesep = .5, ranksep = .25]
    node [fontname = '",def_font,"', fontsize = 14, fontcolor = '",bg_color,"', penwidth=0, color='",ltxt_color,"', style=filled]
    edge [fontname = '",def_font,"', fontcolor = '",ltxt_color,"', color='",ltxt_color,"']
    
    # Input Specs
    Inp [fillcolor = '",ltxt_color,"', label = '(Species s, Term t, Unit u)', shape = rectangle]

    # Conclude Specs
    node [shape = oval]
    Species_Y [label = 'Use the median quantity\nof Species s, Term t and Unit u.', fillcolor = '",get_color(1),"']
    Genus_Y [label = 'Use the median quantity\nof Genus g, Term t and Unit u.', fillcolor = '",get_color(2),"']
    Family_Y [label = 'Use the median quantity\nof Family f, Term t and Unit u.', fillcolor = '",get_color(3),"']
    Order_Y [label = 'Use the median quantity\nof Order o, Term t and Unit u.', fillcolor = '",get_color(4),"']
    Class_Y [label = 'Use the median quantity\nof Class c, Term t and Unit u.', fillcolor = '",get_color(5),"']
    Kingdom_Y [label = 'Use the median quantity\nof Term t and Unit u (across all).', fillcolor = '",txt_color,"']

    # Trigger Specs
    node [shape = diamond, fillcolor = '",bg_color,"', fontcolor = '",txt_color,"', penwidth = 1]
    Species_T [label = 'Does the Species s\nhave >=10 records\nfor the term t and unit u?']
    { rank = same; Species_T, Species_Y }
    Genus_T [label = 'Does g (the Genus of s)\nhave >=10 records\nfor the term t and unit u?']
    { rank = same; Genus_T, Genus_Y }
    Family_T [label = 'Does f (the Family of g)\nhave >=10 records\nfor the term t and unit u?']
    { rank = same; Family_T, Family_Y }
    Order_T [label = 'Does o (the Order of f)\nhave >=10 records\nfor the term t and unit u?']
    { rank = same; Order_T, Order_Y }
    Class_T [label = 'Does c (the Class of o)\nhave any record\nfor the term t and unit u?']
    { rank = same; Class_T, Class_Y }

    # Yes edges
    Inp -> Species_T
    edge [arrowhead = 'box', label = 'Yes']
    Species_T -> Species_Y
    Genus_T -> Genus_Y
    Family_T -> Family_Y
    Order_T -> Order_Y
    Class_T -> Class_Y
    
    # No edges
    edge [label = '     No']
    Species_T -> Genus_T
    Genus_T -> Family_T
    Family_T -> Order_T
    Order_T -> Class_T
    Class_T -> Kingdom_Y

  }
"), height=700)
```

To sum up the flowchart above, we will first count the number of records under the given species, term and unit. If the records are insufficient, we then "roll-up" and attempt to find the number of records for the genus associated with the species. This process continues until we have sufficient data points to calculate the median.

##### Managing Outliers

Another potential challenge of converting around the median is that outliers may be extremely high, thereby skewing the aggregate results towards certain records. To remedy this, we will floor the scaled quantity (\( q_{nt} / m_{nt} \)) of each record to the 5th percentile and cap it to the 95th percentile.

Using the following process, we can now convert all the terms to their respective standardized units.
```{r}
<<pp-term-unit-final>>

post_clean_str()
```

#### Well-Defined Terms

Let us now <span class="hl">convert each term to the universal animal unit</span>. Through observation, the terms can be differentiated into well-defined or ambiguous. A term is well-defined if it is consistent numerically across all animals and records. One good example would be tusk, since 2 of them are always the equivalent of 1 animal. 

Listed below are the terms we describe as well-defined:

```{r}
<<pp-term-well-defined-overview>>

output_str <- "Well-Defined Terms [with Number Per Live Animal]:\n"
wdt <- well_defined_terms %>%
       mutate(outputString = paste0(Term," [", perAnimal, "]"))
output_str <- output_str %>%
              paste0(paste0(wdt$outputString,collapse=", "))
  
cat(output_str)
```

To convert a well-defined term, we simply need to divide the quantity by the number of terms per animal:

```{r}
<<pp-term-well-defined-convert>>

post_clean_str()
```

#### Ambiguous Terms

The final piece of the jigsaw is to convert ambiguous terms. These terms, by nature of their names, are somewhat vague, and there exists no clear association between them and the number of animals. For example, how many ivory pieces make up an elephant? Depending on the sizes of the pieces, the numbers would differ record by record.

To simplify the conversion process, we will revisit our previous assumption when standardizing units of a term. Assuming that <span class="hl">across each species, the median quantity traded in live animal units is the same across any term</span>, ambiguous terms can be converted using the following equation:

\[ q_{animal} = \frac{q_{at}}{m_{at}} \times m_{animal} \]

where <br>
\(q_{animal}\) corresponds to the quantity of animals traded, <br>
\(q_{at}\) corresponds to the quantity traded (in ambiguous terms), <br>
\(m_{at}\) corresponds to the median quantity traded (in ambiguous terms), and <br>
\(m_{animal}\) corresponds to the median quantity of animals traded.

As before, medians are obtained through the <a href="#the-roll-up-median-approach">roll-up approach</a>, and outliers are managed by <a href="#managing-outliers">flooring and capping at the 5th and 95th percentile</a> respectively.

With this final step, we have reduced over 200 distinct term-unit pairs into a single animal unit!

```{r}
<<pp-term-ambiguous-convert>>

post_clean_str()
```

### Completing the Taxonomy

One of the problems with incomplete taxonomies is that identical species have different identities within the data. For example, it is possible that an animal was recorded with its species taxonomy during importing, but only with an Order during exporting. In such circumstances, we might misinterpret the former as being consumed, while the latter being obtained from the local wildlife areas. 

To correct for the incomplete taxonomies, we will be using the method below:

```{r}
grViz(paste0("
  digraph CompleteTax {

    # Default Specs
    graph [compound = true, nodesep = .5, ranksep = .25]
    node [fontname = '",def_font,"', fontsize = 14, fontcolor = '",bg_color,"', penwidth=0, color='",ltxt_color,"', style=filled]
    edge [fontname = '",def_font,"', fontcolor = '",ltxt_color,"', color='",ltxt_color,"']
    
    # Input Specs
    Inp [fillcolor = '",ltxt_color,"', label = 'Incomplete-Taxonomy Trade\nfrom Country A to B\nfor Class c and Order o', shape = rectangle]

    # Conclude Specs
    node [shape = oval]
    Route_Y [label = 'Divide the trades proportionally\namong the Class c, Order o\n trades from A to B.', fillcolor = '",get_color(1),"']
    Export_Y [label = 'Divide the trades proportionally\namong the Class c, Order o\n trades from A.', fillcolor = '",get_color(2),"']
    Import_Y [label = 'Divide the trades proportionally\namong the Class c, Order o\n trades to B.', fillcolor = '",get_color(3),"']
    All_Y [label = 'Divide the trades proportionally\namong all the Class c, Order o\n trades.', fillcolor = '",get_color(4),"']

    # Trigger Specs
    node [shape = diamond, fillcolor = '",bg_color,"', fontcolor = '",txt_color,"', penwidth = 1]
    Route_T [label = 'Does A to B have any taxonomy trades\ncorresponding to Class c and Order o?']
    { rank = same; Route_T, Route_Y }
    Export_T [label = 'Does A have any taxonomy exports\ncorresponding to Class c and Order o?']
    { rank = same; Export_T, Export_Y }
    Import_T [label = 'Does B have any taxonomy imports\ncorresponding to Class c and Order o??']
    { rank = same; Import_T, Import_Y }

    # Yes edges
    Inp -> Route_T
    edge [arrowhead = 'box', label = 'Yes']
    Route_T -> Route_Y
    Export_T -> Export_Y
    Import_T -> Import_Y
    
    # No edges
    edge [label = '     No']
    Route_T -> Export_T
    Export_T -> Import_T
    Import_T -> All_Y

  }
"), height=400)
```

In summary, the methodology above attempts to distribute incomplete taxonomy trades into its species constituents. The approach first analyzes whether there are species of similar natures undergoing the same trade route. If there are, the incomplete taxonomy trades will be distributed amongst the different spcies at existing ratios. If there are none, we subsequently "roll-up" and consider all similar species exported by the country specified under the Exporter column. This process continues until all the incomplete taxonomy trades are mapped to individual species.

### Restricting to Endangered Species

Endangered status of species are located in the IUCN Red List database. Due to licensing restrictions, the data could not be uploaded for public view. However, one is free to create an account and download the csv from <a href="http://www.iucnredlist.org/search/saved?id=90695" target="_blank">here</a>.

By integrating the IUCN data, we can filter non-endangered species out of the dataset.
```{r}
<<pp-endanger>>

post_clean_str()
```


## Exploration

#### Evolution Across Time

The interactive chart below shows how wildlife trade evolves across time:

```{r}
<<exp-time>>

streamgraph_plot
```

Based on the above, we know that as of 2011, species whose statuses are <span class="hl red-text">Critical</span> account for <span class="hl red-text">`r sprintf("%.0f%%",(trades_by_time %>% filter(Year == 2011) %>% mutate(pct = total_trades/sum(total_trades)) %>% filter(IUCNLabel == "Critical"))$pct * 100)`</span> of the total trades. This is disconcerting as the species closest to extinction are the ones being captured the most.

```{r}
pander(
  trades_by_time %>%
    ungroup() %>%
    filter(Year >= 2012) %>%
    group_by(IUCNLabel) %>%
    summarise(MeanTradeGrowth = geoMean(1 + pct_inc) - 1) %>%
    mutate(MeanTradeGrowth = sprintf("%s%.1f%%",ifelse(MeanTradeGrowth > 0,"+","-"),abs(MeanTradeGrowth * 100)))
,caption="Year on Year Trade Growth by Endangered Status")

```

However, when comparing the year-on-year trade growth on average, we see that since 2012, the number of critical-species trading has fallen by an average of <span class="hl red-text">6.1%</span>. A majority of this reduction is due to the fall of trading in the year of 2015. It remains to be seen whether this decrease is a result of improved conservation or due to a lag in reporting.

#### Reasons for Trading

The waffle chart below highlights the purposes behind endangered wildlife trade:

```{r, fig.asp=0.35}
<<exp-purpose>>

waffle_plot
```

Shockingly, trading due to <span class="hl red-text">commercial</span> reasons account for over <span class="hl red-text">95%</span> of all trades! In other words, for every 1 animal traded for species <span class="hl green-text">conservation</span>, another 193 trades are being utilized for profit-making purposes. From the data, the conservation movement is having difficulty catching up with excess capitalism.

#### Most Traded Species

The sunburst plot below allows us to find out which animal class, and which species within each class are traded the most:

```{r}
<<exp-species>>

# Centerize the sunburst plot
htmlwidgets::onRender(
    sunburst_plot,
    "
    function(el, x) {
    $('.sunburst-chart').css('width', 'initial');
    $('.sunburst-chart').css('height', 'initial');
    $('.sunburst-chart').css('left', '50%');
    $('.sunburst-chart').css('transform', 'translate(-50%, 0)');
    }
    "
)

```

By scrolling through the chart above, the <span class="hl yellow-text">Anthozoa</span> class account for close to 70% of the total trades. Some examples of species belonging to this class are soft and hard corals. The well-publicized <span class="hl blue-text">African Elephant</span>, belonging to the <span class="hl blue-text">Mammalia</span> class, interestingly account for less than 0.2% of the total trades.

#### Demand Hotspots

Using export and import data, we can estimate the consumption of wildlife in each country. If a country imports more than it exports, then it is reasonable to assume that the difference is consumed locally.

```{r}
<<exp-imports>>

leaflet_import_plot
```

By utilizing net imports as a proxy, we are able to determine countries with the highest wildlife demand. Surprisingly, <span class="hl red-text">China</span> is ranked fourth, while the <span class="hl red-text">United States</span> has around 5 times more demand than the second-ranked country, <span class="hl red-text">Japan</span>! <span class="hl red-text">Israel</span> and <span class="hl red-text">South Korea</span> make up the top 5.

It remains to be seen whether the huge gap between United States and the other top 4 countries is due to actual trading or higher diligence in reporting. Nevertheless, the map above has surfaced countries (such as Israel and South Korea) that are often overlooked and could be worth focusing more on. 

#### Supply Origins

Using the opposite logic as demand, if a country exports more than it imports, then we assume that the difference is captured within the country.

```{r}
<<exp-exports>>

leaflet_export_plot
```

Using net exports as a proxy, we are able to determine the distribution of supply across the globe. As expected, Southeast Asia and Africa are two areas where the majority of wildlifes are coming from. In particular, <span class="hl purple-text">Indonesia</span> exports the most wildlife, over 2.5 times more than the second-ranked <span class="hl purple-text">Fiji</span>. <span class="hl purple-text">Morocco</span>, the highest-ranked nation in Africa, is 9th.

## Identifying the Major Players

For the purposes of this analysis, the major players in the endangered wildlife trade are countries which take part extensively as either suppliers, consumers or intermediaries. In this section, we will aim to:

1. surface players that were otherwise overlooked, and
2. rank each player according to their wildlife trading activity.

[1] will allow us to have a more macro perspective on the impact of each country towards endangered wildlife, while [2] will help us prioritize which countries to focus on given the limited conservation resources.

### Choosing the Algorithm

The most intuitive way to identify major players is by <span class="hl">their net imports/exports</span>. By ranking these metrics according to size, we can potentially find the countries supplying and consuming endangered wildlife goods. However, one potential pitfall of such an approach is that the intermediary players get overlooked. Consider the following scenario:

```{r}
grViz(paste0("
  digraph expInt {

    # Default Specs
    graph [compound = true, nodesep = .5, ranksep = .25]
    node [fontname = '",def_font,"', fontsize = 14, fontcolor = '",bg_color,"', penwidth=0, color='",ltxt_color,"', style=filled, shape = oval]
    edge [fontname = '",def_font,"', fontcolor = '",ltxt_color,"', color='",ltxt_color,"']
    
    Country_A [label = 'Country\nA', fillcolor = '",get_color(1),"']
    Country_B [label = 'Country\nB', fillcolor = '",get_color(2),"']
    Country_C [label = 'Country\nC', fillcolor = '",get_color(3),"']
    { rank = same; Country_A, Country_B, Country_C }

    Country_A -> Country_B [ label='100' ]
    Country_B -> Country_C [ label='99' ]
    
  }
"), height=50)
```

In such a scenario, <span class="color-1-text">A</span> will have a net export of 100, B a net import of 1 and <span class="color-3-text">C</span> a net import of 99. Using the net import/export approach, <span class="color-2-text">B</span> is the least important player. However, in actuality, <span class="color-2-text">B</span> should be the most pivotal country as it is involved in all 199 trades. Hence, a better approach is required.

One way to tackle the issue above is to instead use each country's gross <span class="hl">imports + exports</span> as a measure of involvement. This approach would work well in the above case, as <span class="color-2-text">B</span> would rightly be identified as the most important player. However, one weakness of such a method would be the huge assumption that the absolute numbers are extremely accurate. In truth, this may not be the case, due to the following reasons:

1. There may be reporting errors recorded in the CITES database, and
2. There could be conversion errors incurred in the <a href="#pre-processing">Pre-Processing</a> stage.

Moreover, consider the following scenario: 

```{r}
grViz(paste0("
  digraph expInt {

    # Default Specs
    graph [compound = true, nodesep = .5, ranksep = .25]
    node [fontname = '",def_font,"', fontsize = 14, fontcolor = '",bg_color,"', penwidth=0, color='",ltxt_color,"', style=filled, shape = oval]
    edge [fontname = '",def_font,"', fontcolor = '",ltxt_color,"', color='",ltxt_color,"']
    
    Country_A [label = 'Country\nA', fillcolor = '",get_color(1),"']
    Country_B [label = 'Country\nB', fillcolor = '",get_color(2),"']
    Country_C [label = 'Country\nC', fillcolor = '",get_color(3),"']
    Country_D [label = 'Country\nD', fillcolor = '",get_color(4),"']
    Country_E [label = 'Country\nE', fillcolor = '",get_color(5),"']

    Country_A -> Country_B [ label='  100' ]
    Country_B -> Country_C [ label='  99' ]
    Country_A -> Country_D [ label='  200' ]
    Country_D -> Country_E [ label='  99' ]
  }
"), height=200)
```

Using the imports + exports approach, we conclude that both <span class="color-3-text">C</span> and <span class="color-5-text">E</span> have the same degree of involvement. However, if the numbers presented are only <span class="hl">roughly</span> accurate, it would be reasonable to assume that <span class="color-3-text">E</span> is more involved as it is linked to a bigger player in <span class="color-2-text">D</span>. (This same argument could be made for drug trafficking, where the closer you are to a drug cartel, the more likely you are to be a major player.)

Fortunately, we are able to resolve these two issues using <a class="hl" href="https://en.wikipedia.org/wiki/PageRank">Google's PageRank</a> algorithm. The algorithm is famous for ranking a website's quality based on:

i. the number of links pointing to the website, and 
ii. who is pointing to the website.

This algorithm can be leveraged for our analysis, which is used to assess the involvement of a country through:

i. the number of trades that country is involved in, and
ii. which other nations that country is involved with.

### Modeling Wildlife Trade

To run the PageRank algorithm, we need to model the wildlife trade as a network, where the countries are represented as <span class="hl">vertices</span>. Two countries are connected by an <span class="hl">edge</span> if the two have wildlife trading activity between one another. The degree of wildlife trading between two countries defines the <span>weight</span> of the edge. For the purposes of this study, we will use the sum of gross imports and exports as an indicator of the intensity of wildlife trading.

Using the above definitions, we are now equipped to model the wildlife trade for the period of 2001 to 2015:

```{r,fig.asp=1.05, fig.width = 8}
<<model-graphs>>
  
edge_bundle_plot
```

The <a href="https://www.r-graph-gallery.com/hierarchical-edge-bundling/" target="_blank">hierarchical edge bundle</a> plot shows the degree of involvement of each country. The larger the circle, the greater the total imports and exports of the country. Unsurprisingly, both <span class="hl blue-text">United States</span> and <span class="hl orange-text">Indonesia</span> have the highest total imports and exports.

Other than that, we can also identify the most popular trade routes. The higher the intensity of wildlife trading, the higher the opacity of that edge. The chart above surfaces some overlooked routes, such as <span class="hl blue-text">Ecuador</span> to <span class="hl orange-text">China</span>. It will be interesting to see how these routes affect the rankings obtained by PageRank.

### The PageRank Algorithm

With the network constructed, we can proceed to run the PageRank algorithm:

```{r}
<<model-pagerank>>

pander(vertices %>% arrange(desc(PRANK)) %>% head(5) %>%
       mutate(Country=NAME, Imports_And_Exports = comma(VALUE), PageRank = percent(PRANK)) %>% 
         select(Country, Imports_And_Exports, PageRank),
       caption="The Top 5 Major Players According to PageRank")
```

Before continuing to the analysis, it is important for us to understand what the PageRank number represents. When used for assessing the quality of websites, this number refers to the <span class="hl">probability of a person viewing the website</span>. The higher the PageRank probability, the more likely the person will view the website, and hence the higher the quality of the site.

Similarly, in the widlife trading context, the PageRank number of a country A is the <span class="hl">probability that A is responsible for a random wildlife trade</span>. The higher the PageRank of a country, the more wildlife trades are attributed to it. In other words, if there are 1 million trades worldwide, <span class="hl">United States</span> would account for approximately <span class="hl">`r round((vertices %>% filter(index == "US"))$PRANK,2)` million</span> trades. By identifying countries with the highest PageRank, we can potentially determine who the major players are.

#### Comparing With Imports + Exports Approach

To comprehend further how PageRank works, let us first compare it with Imports + Exports from a bird's eye view:

```{r, fig.asp=0.5}
<<model-pagerank-hist>>

hist_plot
```

The plot above shows the distribution of <span class="hl blue-text">PageRank</span> and <span class="hl green-text">Imports + Export</span> values. As can be seen, the latter distribution has a much gentler slope, implying that wildlife trade responsibilites are more "spread-out" across countries. <span class="hl">The PageRank algorithm, however, extremizes this distribution by compacting "average" countries into a single peak.</span> Only a handful of least and most important nations are left in the tails, allowing us to better distinguish them.

Secondly, let us investigate how the rankings shift as we transition from one approach to another:

```{r, fig.asp=0.7}
<<model-pagerank-rank>>

rank_plot
```

In both rankings, the top 5 countries remain the same. This implies that <span class="hl">gross imports and exports do play a prominent role in the PageRank algorithm.</span> However, the shifts from top 6 onwards also indicate that total trades is not the only factor influencing PageRank. 

To understand the other variables affecting PageRank, let us compare two countries that have been promoted and demoted extensively, <span class="hl red-text">France</span> and <span class="hl orange-text">South Korea</span>:

```{r, fig.asp=1.05, width=8}
<<model-pagerank-compare>>

compare_plot
```

The diagram above shows all the countries that either France or South Korea traded with. From the diagram, it is clear that <span class="red-text">France</span> has a greater outreach than <span class="orange-text">South Korea</span>, even though the latter has a higher total imports + exports. <span class="hl">This degree of outreach caused the algorithm to promote <span class="red-text">France</span> at the expense of <span class="hl orange-text">South Korea</span>.</span>

#### Decomposing PageRank

% Net Imports times PR, % Net Exports times PR, % Dealed times PR

```{r}
<<model-pagerank-decompose>>

```

### The Major Players

Listed below are the major players detected through the PageRank algorithm. These players have been divided into three groups (<span class="hl">Consumers</span>, <span class="hl">Dealers</span> and <span class="hl">Suppliers</span>) based on their import percentage (with respect to the total imports and exports).

```{r}
<<model-results>>

pander(players, caption="The Top Suppliers, Dealers and Consumers")
```

```{r echo=FALSE, warning=FALSE, results='hide'}
<<results>>
```

#### Consumers

Consumers are countries that demand wildlife goods. They fall in the top 40th percentile of import percentages.

Please scroll through the icons below to find out more about the top 5 consumers:

`r get_players((consumers %>% head(5))$index,get_color("red"), type="Net_Imports")`

#### Dealers

Dealers are countries that neither consume nor supply wildlife goods. They act as intermediaries between suppliers and consumers and fall between the 40th to 60th percentiles of import percentages.

Please scroll through the icons below to find out more about the top 5 dealers:

`r get_players((dealers %>% head(5))$index,heading_color, type="Transits")`

#### Suppliers

Suppliers are countries that provide for wildlife goods. They fall in the bottom 40th percentile of import percentages.

Please scroll through the icons below to find out more about the top 5 suppliers:

`r get_players((suppliers %>% head(5))$index,get_color("purple"), type="Net_Exports")`

## Summary of Results

By standardizing measurements through <a href="#the-roll-up-median-approach">the roll-up median approach</a> and ranking countries via <a href="#the-pagerank-algorithm">the PageRank algorithm</a>, we have identified the major players in endangered wildlife trade:

* <span class="hl red-text">Consumers</span> are countries which generate the demand of wildlife. It is important to educate citizens in these countries to stop using wildlife-related products. The top 5 consumers are `r paste0(sprintf('<span class="hl red-text">%s</span>',(consumers %>% head(5))$NAME), collapse = ", ")`.
* <span class="hl">Dealers</span> are countries which connect suppliers with consumers. More stringent measures can be set up in these countries' trading ports to reduce endangered wildlife trading. The top 5 dealers are `r paste0(sprintf('<span class="hl">%s</span>',(dealers %>% head(5))$NAME), collapse = ", ")`.
* <span class="hl purple-text">Suppliers</span> are countries that provide the wildlife goods. Efforts to protect the nature reserves of these countries should be emphasized. The top 5 suppliers are `r paste0(sprintf('<span class="hl purple-text">%s</span>',(suppliers %>% head(5))$NAME), collapse = ", ")`.

## Limitations

* Completing Taxonomy -> Problematic as some of these might not be endangered
* Wild
* Unidentified Species
* Overview

## References

